---
title: "`EpiCompare`: QC and Benchmarking of Epigenetic Datasets"
author: "Sera Choi"
date: "23/11/2021"
output: 
  github_document:
editor_options:
  chunk_output_type: inline
---

# Introduction 

`EpiCompare` is an R package for QC and benchmarking epigenetic datasets. 
It currently performs two functions:

1. Calculates percentage of overlapping peaks in two peak files 
2. Performs ChromHMM for individual peak files, overlapping and unique peaks. 

The function outputs a report in HTML format. 

# Installation 

How to install the package. 

# Usage 

Load package and example datasets. The two example peakfiles have been 
```{r setup}
library(EpiCompare)
data("encode_H3K27ac")
data("CnT_H3K27ac")
```

The two example peakfiles have been transformed into GRanges object. 
If your peak files are in BED format, simply use `ChIPseeker::readPeakFile("/path/to/peak/file", as = "GRanges")`.

```{r, eval=FALSE}
EpiCompare(peakfile1 = encode_H3K27ac,
           peakfile1_name = "encode",
           peakfile2 = CnT_H3K27ac,
           peakfile2_name = "CUT&Tag",
           outpath = "./EpiCompare.html")
```


# Session Info

<details>

```{r Session Info}
utils::sessionInfo()
```

</details>














