---
title: "EpiCompare Report"
author: "Sera Choi"
output: 
  html_document: 
    toc: yes
    toc_float: yes
params:
  peakfile: 
    value: NULL
  names: 
    value: NULL
  output_dir: 
    value: "./"
  blacklist: 
    value: NULL
  picard_list:
    value : NULL
  picard_names:
    value : NULL
  reference: 
    value: NULL
  stat_plot:
    value: FALSE
  chrmHMM_plot:
    value: FALSE
  save_output:
    value: FALSE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
if(params$save_output){
  outfile_dir <- paste0(params$output_dir,"/EpiCompare_file")
  if(!dir.exists(outfile_dir)){
    dir.create(outfile_dir)
  }
}
```

EpiCompare compares different epigenetic datasets for benchmarking and quality control purposes. 
The report consists of three main sections:

1. **General Metrics:** Metrics on fragments (duplication rate) and peaks (blacklisted peaks and widths)
2. **Peak Overlaps:** Percentage and statistical significance of overlapping and unique peaks
3. **Functional Annotation:** Functional annotation (ChromHMM, motif and enrich) of peaks. 

Input peak files. Total of `r length(params$names)` samples: 

```{r name, echo=FALSE}
n <- 1
for (name in params$names){
  print(paste0("File", n, ": ", name))
  n <- n + 1
}
```

<br>

## 1. General Metrics {.tabset}

<br>

### Fragment Information

This information is displayed only if summary metrics from Picard is provided. See help manual. 

* Mapped_Fragments: Number of mapped read pairs in the file. 
* Duplication_Rate: Percentage of mapped sequence that is marked as duplicate.
* Unique_Fragments: Number of mapped sequence that is not marked as duplicate. 

<br>

```{r, echo=FALSE, warning=FALSE}
if (!is.null(params$picard_list)){
  fragment_info_df <- fragment_info(picard_list = params$picard_list,
                                    namelist = params$picard_names)
  knitr::kable(fragment_info_df)
  
  if(params$save_output){
    write.table(fragment_info_df, file=paste0(outfile_dir,"/fragment_info"), sep="\t")
    }
}   
```

### Peak Information 

* Total_N: Total number of peaks including those blacklisted. 
* Blacklisted_Peaks: Percentage of blacklisted peaks present in the sample. 

<br>

```{r, echo=FALSE, warning=FALSE}
peak_info_df <- peak_info(peak_list = params$peakfile,
                          file_names = params$names,
                          blacklist = params$blacklist)
knitr::kable(peak_info_df)

if(params$save_output){
  write.table(peak_info_df, file=paste0(outfile_dir,"/peak_info"), sep="\t")
}
```

### Peak widths 

Distribution of peak widths in each sample after removing blacklisted peaks. 

<br> 

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# remove blacklist regions 
peaklist_blrm <- list()
for (sample in params$peakfile){
  peak_blacklist_removed <- IRanges::subsetByOverlaps(sample, params$blacklist, invert = TRUE)
  peaklist_blrm <- c(peaklist_blrm, peak_blacklist_removed)
}

width_plot <- width_boxplot(peaklist = peaklist_blrm,
                            namelist = params$names)
print(width_plot)

if(params$save_output){
  ggplot2::ggsave(filename = "width_plot.png",plot = width_plot, device = "png", path = outfile_dir)
  unlink("width_plot.png")
}
```

<br>

## 2. Peak Overlaps {.tabset}

### Individual samples 

Heatmap of percentage overlaps between input peak files. 
Hover over the heatmap for percentage values. 
<br>

```{r, echo=FALSE, fig.width=7, warning=FALSE}

overlap_heatmap <- overlap_heatmap(peaklist = peaklist_blrm,
                  namelist = params$names,
                  save_dir = NULL)
overlap_heatmap

if(params$save_output){
  overlap_heatmap(peaklist = peaklist_blrm,
                  namelist = params$names,
                  save_dir = paste0(outfile_dir,"/samples_percent_overlap.html"))
  unlink(paste0(outfile_dir,"/samples_percent_overlap_files"), recursive = TRUE)
  }

```

### Individual samples + other histone modifications

Heatmap of percentage overlaps between input peak files and other histone marks from ENCODE. 
Hover over the heatmap for percentage values. 
<br>

```{r, echo=FALSE, fig.width=7, warning=FALSE}
overlap_heatmap <- overlap_heatmap(peaklist = peaklist_blrm,
                  namelist = params$names,
                  save_dir = NULL)
overlap_heatmap

if(params$save_output){
  overlap_heatmap(peaklist = peaklist_blrm,
                  namelist = params$names,
                  save_dir = paste0(outfile_dir,"/samples_percent_overlap.html"))
  unlink(paste0(outfile_dir,"/samples_percent_overlap_files"), recursive = TRUE)
  }
```

### Significance of overlapping vs unique peaks 

The plot is displayed only if a reference peak file is provided and `stat_plot = TRUE`. 
Depending on the format of the reference file, the function output different plots:

* If the reference file has BED6+4 format (peak called with MACS2), the plot is a paired boxplot showing a distribution 
of -log10(q-value) for overlapping and unique peaks per sample.
* If the reference file does not have BED6+4 format, it generates a barplot of percentage overlap per sample, coloured by adjusted p-value.

<br>

```{r, echo=FALSE, fig.width=7, warning=FALSE, message=FALSE}
if (is.null(params$reference) == FALSE){
  if (params$stat_plot){
    stat_plot <- overlap_stat_plot(reference = params$reference,
                                   peaklist = peaklist_blrm,
                                   namelist = params$names)
    print(stat_plot[[1]])
    
    if(params$save_output){
      ggplot2::ggsave(filename = "stat_plot.png", plot = stat_plot[[1]], device = "png", path = outfile_dir)
      unlink("stat_plot.png")
      }
  }
}
```


## 3. Functional Annotation 

### 3.1 ChromHMM {.tabset}

ChromHMM annotates different chromatin states [(ChromHMM)](http://compbio.mit.edu/ChromHMM/#:~:text=ChromHMM%20is%20software%20for%20learning%20and%20characterizing%20chromatin%20states.&text=ChromHMM%20is%20based%20on%20a,one%20or%20more%20cell%20types.). The annotation used were obtained from [here](http://hgdownload.cse.ucsc.edu/goldenPath/hg19/encodeDCC/wgEncodeBroadHmm/).   


#### All samples 

ChromHMM annotation of individual peak files. 

```{r, echo=FALSE, fig.width=7, warning=FALSE, message=FALSE}
if (params$chrmHMM_plot){
  chrHMM_url = "http://hgdownload.cse.ucsc.edu/goldenPath/hg19/encodeDCC/wgEncodeBroadHmm/wgEncodeBroadHmmK562HMM.bed.gz"
  chrHMM = genomation::readBed(chrHMM_url)
  chrHMM_list = GenomicRanges::split(chrHMM, chrHMM$name, drop = TRUE) 
  
  samples_chrmHMM <- plot_chrmHMM(peaklist = peaklist_blrm,
                                  namelist = params$names,
                                  chrmHMM_annotation = chrHMM_list)
  print(samples_chrmHMM)
  
  if(params$save_output){
      ggplot2::ggsave(filename = "samples_chromHMM.png", plot = samples_chrmHMM, device = "png", path = outfile_dir)
      unlink("samples_chromHMM.png")
  }
}
```

#### Overlapping peaks 

ChromHMM annotation of peaks overlapping with the reference peak file. 

```{r, echo=FALSE, fig.width=7, warning=FALSE, message=FALSE}
if (params$chrmHMM_plot){
  if (is.null(params$reference)==FALSE){
    overlap_list <- list()
    
    for(file in peaklist_blrm){
      overlap <- IRanges::subsetByOverlaps(x = file, ranges = params$reference)
      overlap_list <- c(overlap_list, overlap)
      }
    
    overlap_chrmHMM <- plot_chrmHMM(peaklist = overlap_list,
                                    namelist = params$names,
                                    chrmHMM_annotation = chrHMM_list)
    
    print(overlap_chrmHMM)

    if(params$save_output){
      ggplot2::ggsave(filename = "overlap_chromHMM.png", plot = overlap_chrmHMM, device = "png", path = outfile_dir)
      unlink("overlap_chromHMM.png")
    }
  }
}
```


#### Unique peaks 

ChromHMM annotation of peaks non-overlapping with the reference peak file. 

```{r, echo=FALSE, fig.width=7, warning=FALSE, message=FALSE}

if (params$chrmHMM_plot){
  if (is.null(params$reference)==FALSE){
      unique_list <- list()
    
  for(file in peaklist_blrm){
    unique <- IRanges::subsetByOverlaps(x = file, ranges = params$reference, invert = TRUE)
    unique_list <- c(unique_list, unique)
    }
    
    unique_chrmHMM <- plot_chrmHMM(peaklist = unique_list,
                                   namelist = params$names,
                                   chrmHMM_annotation = chrHMM_list)
    print(unique_chrmHMM)

    if(params$save_output){
      ggplot2::ggsave(filename = "unique_chromHMM.png", plot = unique_chrmHMM, device = "png", path = outfile_dir)
      unlink("unique_chromHMM.png")
    }
  }
}
```

### 3.2 Enrichment analysis 

### 3.3 Motif analysis
